---
title: "Project 1"
author: "Shiyue Mao"
date: "1/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Load data to R.

```{r, echo=TRUE}
library(readr)
activity <- read_csv("activity.csv")
```

## Mean total number of steps per day

A quick glance at the dataset and the first 6 observations.
``` {r, echo=TRUE}
summary(activity)
head(activity)
```

Calculate the total number of steps by introducing a new library dplyr, and then plot a histogram about it. After that, also show mean and median number for total steps per day.
```{r, echo=TRUE}
library(dplyr)
step_by_date <- aggregate(steps~date, data = activity, FUN = sum, na.rm = TRUE)
colnames(step_by_date) <- c("date", "total_steps")
# show total steps
step_by_date

# make a histogram plot
hist(step_by_date$total_steps, breaks = seq(0,25000, by = 2500), xlab = "Total steps per day", main = "Total steps taken per day")

# take a look for mean and median value of the statistics
paste("The mean is ", mean(step_by_date$total_steps))
paste("The median is ",median(step_by_date$total_steps))
```

## Average daily activity pattern

Firstly, make a time series plot with 5 min interval and average steps taken each day throughout the whole period.

```{r, echo=TRUE}
avg_activity <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
names(avg_activity) <- c("interval", "steps")
with(avg_activity, plot(steps~interval, type = "l"))
```

Then, figure out which 5 min interval contains the maximum number of steps.
```{r, echo=TRUE}
avg_activity[which.max(avg_activity$steps),]$interval
```

## Input missing values

Firstly, search total number of missing values (NAs) in the dataset. It seems there are 2304 NAs in the dataset.
```{r, echo=TRUE}
sum(is.na(activity))
```

Secondly, filling the NAs by mean of 5 min interval, and create a new dataset named new_activity

```{r, echo=TRUE}
activity$steps[is.na(activity$steps)] <- avg_activity$steps
new_activity <- activity
```

Then, make a histogram again with mean and median values.

```{r, echo=TRUE}
new_step_by_date <- aggregate(steps~date, new_activity, FUN = sum)
colnames(new_step_by_date) <- c("date", "steps")
hist(new_step_by_date$steps, breaks = seq(0,25000, by = 2500), xlab = "Total steps per day (new)", main = "Total steps taken per day (new)")
paste("The new mean is ", mean(new_step_by_date$steps))
paste("The new median is ",median(new_step_by_date$steps))
```
After being compared with the original dataset, it seems that filling of data with mean value of intervals increases the estimation of total daily number of steps.

## Are there differences in activity patterns between weekdays and weekends?

Determine if the day is weekday or weekend.
```{r, echo=TRUE}
new_activity$weekday <- sapply(new_activity$date, function(x){
  if (weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
      {y <- "Weekend"} 
  else
      {y <- "Weekday"}
      y
})
```

make a new panel plot.
```{r, echo=TRUE}
library(ggplot2)
new_activity_by_date <- aggregate(steps~interval + weekday, data = new_activity, FUN = mean, na.rm = TRUE)
ggplot(new_activity_by_date, aes(x = interval, y = steps, colour = weekday)) +
  geom_line() +
  facet_wrap(~weekday, ncol = 1, nrow = 2)
```